Stellar Radiation is Critical for Regulating Star Formation and Driving Outflows in Low Mass Dwarf Galaxies

### 2019-03-08 thesis/chapters/ch_chapter2.tex
first commit of radiation paper

\documentclass[twocolumn]{aastex62}

\newcommand{\vdag}{(v)^\dagger}
\newcommand\aastex{AAS\TeX}
\newcommand\latex{La\TeX}

\graphicspath{{./}{figures/}}
\usepackage{natbib}
\bibliographystyle{apj}

\submitjournal{ApJ}

\shorttitle{Radiation Feedback in Dwarf Galaxies}
\shortauthors{Emerick et. al.}

\begin{document}

\title{Stellar Radiation is Critical for Regulating Star Formation and Driving Outflows in Low Mass Dwarf Galaxies}

\correspondingauthor{Andrew Emerick}
\email{emerick@astro.columbia.edu}

\author{Andrew Emerick}
\affil{Department of Astronomy, Columbia University, New York, NY, 10027, USA}
\affil{Department of Astrophysics, American Museum of Natural History, New York, NY 10024}

\author{Greg L. Bryan}
\affiliation{Department of Astronomy, Columbia University, New York, NY, 10027, USA}
\affiliation{Center for Computational Astrophysics, Flatiron Institute, 162 5th Ave, New York, NY, 10003, U.S.A}

\author{Mordecai-Mark Mac Low}
\affil{Department of Astrophysics, American Museum of Natural History, New York, NY 10024}
\affiliation{Department of Astronomy, Columbia University, New York, NY, 10027, USA}

\nocollaboration

% Letter word count:
%    Text (including captions) : 3500
%    Abstract                  : 500
%    # Figures                 : 5
%       max panels per fig     : 9
%    # references              : 50

\begin{abstract}
%
%
% NOTE: Overleaf includes abstract words in word count in text... subtract
%       this from the total.
%
%
Effective stellar feedback is used in models of galaxy formation to drive realistic galaxy evolution. Models typically include energy injection from supernovae as the dominant form of stellar feedback, often in some form of sub-grid recipe. However, it has been recently suggested that pre-SN feedback (stellar winds or radiation) is necessary in high-resolution simulations of galaxy evolution to properly regulate star formation and properties of the interstellar medium (ISM). Following these processes is computationally challenging, so many prescriptions model this feedback approximately, accounting for the local destruction of dense gas clouds around newly formed stars in lieu of a full radiative transfer calculation. In this work we examine high resolution simulations (1.8~pc) of an isolated dwarf galaxy with detailed stellar feedback tracked on a star-by-star basis. By following stellar ionizing radiation with an adaptive ray-tracing radiative transfer method, we test its importance in regulating star formation and driving outflows in this galaxy. We find that including ionizing radiation reduces the star formation rate (SFR) by over a factor of 5, and is necessary to produce the ISM conditions needed for supernovae to drive significant outflows. We find that a localized approximation for radiation feedback is sufficient to regulate the SFR on short timescales, but does not allow significant outflows. Short and long range radiation effects are both important in driving the evolution of our low-metallicity, low-mass dwarf galaxy. Generalizing these results to more massive galaxies would be a valuable avenue of future research.
\end{abstract}

%% Keywords should appear after the \end{abstract} command. 
%% See the online documentation for the full list of available subject
%% keywords and the rules for their use.
\keywords{galaxies -- feedback -- galaxy evolution}

@@

In this work we use a detailed model for stellar feedback presented in \cite{Emerick2018} (hereafter Paper I) to study the role of radiation feedback in dwarf galaxy evolution. For the first time in a galaxy-scale simulation, we resolve individual HII regions using an adaptive ray tracing radiative transfer method to follow the ionizing radiation from particles that represent individual stars. We focus on addressing two questions: 1) what role does radiation feedback play in regulating star formation, and 2) are the long-range effects of radiation feedback important, or is the local destruction of dense gas the dominant effect.

@@

We include a UV background, tabulated metal line cooling, and a 9 species non-equilibrium chemistry solver using \textsc{Grackle} \citep{GrackleMethod}. Star formation occurs stochastically in cold, dense regions ($n > 200$~cm$^{-3}$, $T < 200$~K) by sampling a \cite{Salpeter1955} IMF from 1~M$_{\odot}$ to 100~M$_{\odot}$ and depositing \textit{individual} star particles over this mass range. We include feedback from stellar winds, AGB winds, FUV and LW band radiation which drives photoelectric heating and H$_2$ dissociation respectively, HI and HeI ionizing radiation, and core collapse and Type Ia SNe using thermal energy injection. FUV and LW band radiation are both taken to be optically thin, with local (cell-by-cell) attenuation alone. Ionizing radiation is followed using radiative transfer, as discussed below. Stellar lifetimes, surface gravity, effective temperature, and radii are set by the initial stellar mass and metallicity through interpolation over the PARSEC \citep{Bressan2012} zero age main sequence values. These properties are used to set the FUV, LW, and ionizing photon fluxes from each star through interpolation on the OSTAR2002 \citep{LanzHubeny2003} grid. We only model the radiation from stars with masses above 8~M$_{\odot}$.

@@

\begin{figure*}
\centering
\includegraphics[width=0.49\linewidth]{sfr}
\includegraphics[width=0.49\linewidth]{mass}
\caption{The star formation rate (left) and gas and stellar mass (right) of each of our three simulations over time. Time bins with no star formation are left empty.}
\label{fig:sfr_mass_evolution}
\end{figure*}

@@

\begin{figure}
\centering
\includegraphics[width=0.99\linewidth]{mass_density_cut}
\caption{The total gas mass at $n > 200$~cm$^{-3}$, our star formation density threshold.}
\label{fig:sf gas}
\end{figure}

@@

\begin{figure*}
\centering
\includegraphics[width=0.32\linewidth]{metal_retention}
\includegraphics[width=0.32\linewidth]{mass_outflow_comparison} 
\includegraphics[width=0.32\linewidth]{mass_loading_comparison}
\caption{{\bf Left:} The fraction of metals contained within the ISM of each galaxy (blue) and the fraction ejected beyond the virial radius (orange). As in Figure~\ref{fig:sfr_mass_evolution}, the fiducial run is given as solid lines, while the noRT and shortrad are dash-dotted and dashed respectively. {\bf Middle:} The mass outflow rate for each run at 0.25~R$_{\rm vir}$ (solid) and R$_{\rm vir}$ (dashed). {\bf Right:} The mass loading factor $\eta$ calculated as the outflow rate divided by the SFR averaged over 100~Myr. Inclusion of diffuse ionization makes an order of magnitude difference in $\eta$ in this case. }
\label{fig:metal_retention}
\end{figure*}

@@

\begin{figure*}
\centering
\includegraphics[width=0.99\linewidth]{DD0136_fiducial_shortrad_nort}
\caption{Edge-on slices of each simulation showing gas number density (left), temperature (middle), and hydrogen ionization fraction (right) 17~Myr after the formation of the first star in each run. Each panel is 4 kpc x 4 kpc. The full movie of this evolution \textbf{from 0 to 118 Myr} is included for further clarification. \textbf{This movie demonstrates that the differences in outflows between the simulations depends on the ability for feedback to carve low-density, ionized channels along which SNe can launch gas beyond the disk of the galaxy. These channels are readily created by ionizing radiation in the fiducial simulation, or to some degree by SNe in the noRT simulation due to its much higher SN rate. However, SNe launched gas is trapped by neutral, cold shells in the shortrad simulation, unable to easily escape from the disk of the simulated galaxy. These same insights can also be gained by comparing this figure with Figure~\ref{panel2}.}}
\label{fig:panel1}}
\label{fig:panel1}
\end{figure*}

\begin{figure*}
\centering
\includegraphics[width=0.99\linewidth]{DD0160_fiducial_shortrad_nort}
\caption{Same as Figure~\ref{fig:panel1}, but at 40~Myr.}
\label{fig:panel2}
\end{figure*}

@@

\acknowledgments
A.E. is funded by the NSF Graduate Research Fellowship DGE 16-44869. G.L.B. acknowledges support from NSF AST-1312888, NASA NNX15AB20G, and NSF AST-1615955. M.-M.M.L. was partly funded by NASA  grant NNX14AP27G and NSF grant AST18-15461. We gratefully recognize computational resources provided by NSF XSEDE through grant number TGMCA99S024, the NASA High-End Computing Program through the NASA Advanced Supercomputing Division at Ames Research Center, Columbia University, and the Flatiron Institute. This work made significant use of many open source software packages, including \textsc{yt}, \textsc{Enzo}, \textsc{Grackle}, \textsc{Python}, \textsc{IPython}, \textsc{NumPy}, \textsc{SciPy}, \textsc{Matplotlib}, \textsc{HDF5}, \textsc{h5py}, \textsc{Astropy}, \textsc{Cloudy} and \textsc{deepdish}. These are products of collaborative effort by many independent developers from numerous institutions around the world. Their commitment to open science has helped make this work possible.

\software{astropy \citep{astropy} matplotlib \citep{matplotlib}, Numpy \citep{numpy}, SciPy \citep{scipy}, IPython \citep{Ipython}, yt \citep{yt}, Enzo \citep{Enzo2014}, Grackle \citep{GrackleMethod}, CLOUDY \citep{cloudy}, HDF5 \citep{HDF5}}

\bibliography{msbib}

\end{document}

>>>>

\chapter[Stellar Radiation is Critical for Regulating Star Formation and Driving Outflows in Low Mass Dwarf Galaxies]{Stellar Radiation is Critical for Regulating Star Formation and Driving Outflows in Low Mass Dwarf Galaxies \label{ch:chapter2}}
\let\thefootnote\relax\footnotetext{This section contains text from an article published originally as \cite{Emerick2018a}}

%
% Copy paste paper here, without abstract or keywords
%
%\begin{abstract}
%
%Effective stellar feedback is used in models of galaxy formation to drive realistic galaxy evolution. Models typically include energy injection from supernovae as the dominant form of stellar feedback, often in some form of sub-grid recipe. However, it has been recently suggested that pre-SN feedback (stellar winds or radiation) is necessary in high-resolution simulations of galaxy evolution to properly regulate star formation and properties of the interstellar medium (ISM). Following these processes is computationally challenging, so many prescriptions model this feedback approximately, accounting for the local destruction of dense gas clouds around newly formed stars in lieu of a full radiative transfer calculation. In this work we examine high resolution simulations (1.8~pc) of an isolated dwarf galaxy with detailed stellar feedback tracked on a star-by-star basis. By following stellar ionizing radiation with an adaptive ray-tracing radiative transfer method, we test its importance in regulating star formation and driving outflows in this galaxy. We find that including ionizing radiation reduces the star formation rate (SFR) by over a factor of 5, and is necessary to produce the ISM conditions needed for supernovae to drive significant outflows. We find that a localized approximation for radiation feedback is sufficient to regulate the SFR on short timescales, but does not allow significant outflows. Short and long range radiation effects are both important in driving the evolution of our low-metallicity, low-mass dwarf galaxy. Generalizing these results to more massive galaxies would be a valuable avenue of future research.
%\end{abstract}

@@

In this work we use a detailed model for stellar feedback presented in \cite{Emerick2019} (hereafter Paper I) to study the role of radiation feedback in dwarf galaxy evolution. For the first time in a galaxy-scale simulation, we resolve individual HII regions using an adaptive ray tracing radiative transfer method to follow the ionizing radiation from particles that represent individual stars. We focus on addressing two questions: 1) what role does radiation feedback play in regulating star formation, and 2) are the long-range effects of radiation feedback important, or is the local destruction of dense gas the dominant effect.

@@

We include a UV background, tabulated metal line cooling, and a 9 species non-equilibrium chemistry solver using \textsc{Grackle} \citep{GrackleMethod}. Star formation occurs stochastically in cold, dense regions ($n > 200$~cm$^{-3}$, $T < 200$~K) by sampling a \cite{Salpeter1955} IMF from 1~M$_{\odot}$ to 100~M$_{\odot}$ and depositing \textit{individual} star particles over this mass range. We include feedback from stellar winds, AGB winds, FUV and LW band radiation which drives photoelectric heating and H$_2$ dissociation respectively, HI and HeI ionizing radiation, and core collapse and Type Ia SNe using thermal energy injection. FUV and LW band radiation are both taken to be optically thin, with local (cell-by-cell) attenuation alone. Ionizing radiation is followed using radiative transfer, as discussed below. Stellar lifetimes, surface gravity, effective temperature, and radii are set by the initial stellar mass and metallicity through interpolation over the PARSEC \citep{Bressan2012} zero age main sequence values. These properties are used to set the FUV, LW, and ionizing photon fluxes from each star through interpolation on the OSTAR2002 \citep{Lanz2003} grid. We only model the radiation from stars with masses above 8~M$_{\odot}$.

@@

\begin{figure*}
\centering
\includegraphics[width=0.49\linewidth]{figures/ch2/sfr}
\includegraphics[width=0.49\linewidth]{figures/ch2/mass}
\caption{The star formation rate (left) and gas and stellar mass (right) of each of our three simulations over time. Time bins with no star formation are left empty.}
\label{fig:sfr_mass_evolution}
\end{figure*}

@@

\begin{figure}
\centering
\includegraphics[width=0.99\linewidth]{figures/ch2/mass_density_cut}
\caption{The total gas mass at $n > 200$~cm$^{-3}$, our star formation density threshold.}
\label{fig:sf gas}
\end{figure}

@@

\begin{figure*}
\centering
\includegraphics[width=0.32\linewidth]{figures/ch2/metal_retention}
\includegraphics[width=0.32\linewidth]{figures/ch2/mass_outflow_comparison} 
\includegraphics[width=0.32\linewidth]{figures/ch2/mass_loading_comparison}
\caption{{\bf Left:} The fraction of metals contained within the ISM of each galaxy (blue) and the fraction ejected beyond the virial radius (orange). As in Figure~\ref{fig:sfr_mass_evolution}, the fiducial run is given as solid lines, while the noRT and shortrad are dash-dotted and dashed respectively. {\bf Middle:} The mass outflow rate for each run at 0.25~R$_{\rm vir}$ (solid) and R$_{\rm vir}$ (dashed). {\bf Right:} The mass loading factor $\eta$ calculated as the outflow rate divided by the SFR averaged over 100~Myr. Inclusion of diffuse ionization makes an order of magnitude difference in $\eta$ in this case. }
\label{fig:metal_retention}
\end{figure*}

@@

\begin{figure*}
\centering
\includegraphics[width=0.99\linewidth]{figures/ch2/DD0136_fiducial_shortrad_nort}
\caption{Edge-on slices of each simulation showing gas number density (left), temperature (middle), and hydrogen ionization fraction (right) 17~Myr after the formation of the first star in each run. Each panel is 4 kpc x 4 kpc. The full movie of this evolution from 0 to 118 Myr is included for further clarification. This movie demonstrates that the differences in outflows between the simulations depends on the ability for feedback to carve low-density, ionized channels along which SNe can launch gas beyond the disk of the galaxy. These channels are readily created by ionizing radiation in the fiducial simulation, or to some degree by SNe in the noRT simulation due to its much higher SN rate. However, SNe launched gas is trapped by neutral, cold shells in the shortrad simulation, unable to easily escape from the disk of the simulated galaxy. These same insights can also be gained by comparing this figure with Figure~\ref{fig:panel2}.}
\label{fig:panel1}
\end{figure*}

\begin{figure*}
\centering
\includegraphics[width=0.99\linewidth]{figures/ch2/DD0160_fiducial_shortrad_nort}
\caption{Same as Figure~\ref{fig:panel1}, but at 40~Myr.}
\label{fig:panel2}
\end{figure*}

@@

\section*{Acknowledgments:} 
A.E. is funded by the NSF Graduate Research Fellowship DGE 16-44869. G.L.B. acknowledges support from NSF AST-1312888, NASA NNX15AB20G, and NSF AST-1615955. M.-M.M.L. was partly funded by NASA  grant NNX14AP27G and NSF grant AST18-15461. We gratefully recognize computational resources provided by NSF XSEDE through grant number TGMCA99S024, the NASA High-End Computing Program through the NASA Advanced Supercomputing Division at Ames Research Center, Columbia University, and the Flatiron Institute. This work made significant use of many open source software packages, including \textsc{yt}, \textsc{Enzo}, \textsc{Grackle}, \textsc{Python}, \textsc{IPython}, \textsc{NumPy}, \textsc{SciPy}, \textsc{Matplotlib}, \textsc{HDF5}, \textsc{h5py}, \textsc{Astropy}, \textsc{Cloudy} and \textsc{deepdish}. These are products of collaborative effort by many independent developers from numerous institutions around the world. Their commitment to open science has helped make this work possible.

%\software{astropy \citep{astropy} matplotlib \citep{matplotlib}, Numpy \citep{numpy}, SciPy \citep{scipy}, IPython \citep{Ipython}, yt \citep{yt}, Enzo \citep{Enzo2014}, Grackle \citep{GrackleMethod}, CLOUDY \citep{cloudy}, HDF5 \citep{HDF5}}

%************* APPENDICES ************************%

%\clearpage
%\appendix
\setcounter{section}{0}%
\renewcommand\thesection{\thechapter.\Alph{section}}

\renewcommand\thesection{\thechapter.\arabic{section}}
